---
title: "Can we have it all?"
author: "Philip Unger & Philipp Staender"
date: "2 May 2016"
output:
  ioslides_presentation:
    widescreen: true
    number_sections: yes
    fig_caption: yes
    keep_md: true
    background: white
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load packages and create BibTeX file for R-packages
PackagesUsed <- c("ggplot2", "repmis", "plyr", "dplyr", "MASS", "Hmisc", "interplot", "gridExtra", "repmis", "car", "reshape", "gmodels", "quantmod", "foreign")

# Set working directory
#setwd("~/Documents/CDA/collaborative_projects/Final_Project_P-P") # Unger
setwd("~/R_data/Final_Project_P-P") # Staender

# Load PackagesUsed
repmis::LoadandCite(PackagesUsed, file = "packages_pres.bib", install = FALSE)
```

```{r, include=FALSE}
if(!file.exists("data/data_final.rda")) {
  source('data_combine.R')
}

# Opens data_final.rda if not already loaded as an object
if(!exists("z.df" )) {
  load("data/data_final.rda")
}
```

## Intro

<div class="black">
Research Question
</div>

*"Are career pursuits reconcilable with a happy life?"* 

<div class="black">
The debate
</div>

*'I still strongly believe that women can “have it all” (and that men can too). ... But not today, not with the way America’s economy and society are currently structured.'*

<div class="black">
Anne-Marie Slaughter - Former director of policy planning for the U.S. State Department
</div>


## Research Design 

### Data

<div class="black">
General Social Survey
</div>

- Cross-sectional survey of the adult population in the United States
- Conducted between 1972 to 2014
- Sample size around 60,000

<div class="black">
Current Population Survey
</div>

- Used to generate income percentiles in age and educational groups (reference group income)

<div class="black">
Personal Consumption Expenditure (price-deflator)
</div>

## Research Design 

<div class="black">
Methodology
</div>

- Graphical analysis
- Linear Probability Model

<div class="black">
Operationalization
</div>

- Defining career
- Subjective well being measures

# Descriptive statistics

## Measures of happiness

```{r,warning=FALSE,echo=FALSE,fig.cap="Figure 1: Distribution of reported happiness, job-satisfaction and happiness with marriage"}

z$hapmarr <- NA
z$hapmarr[z$hapmar==3] <- 1
z$hapmarr[z$hapmar==2] <- 2
z$hapmarr[z$hapmar==1] <- 3

z$hapmar <- NULL
z <- rename(z, c(hapmarr="hapmar"))

happy_dist <- table(z$happy)/sum(table(z$happy))*100
satjob_dist <- table(z$satjob)/sum(table(z$happy))*100
t <- subset(z, z$married == 1)
hapmar_dist <- table(t$hapmar)/sum(table(t$happy))*100

par(mfrow=c(1,3)) 

barplot(happy_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness",
        ylab = "Frequency (percent)",
        legend = c("Happiness"),
        ylim = c(0, 80))  

barplot(satjob_dist, 
        col=c("navyblue"),
        xlab = "Reported job-satisfaction",
        legend = c("Job-satisfcation"),
        ylim = c(0, 80))

barplot(hapmar_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness with marriage",
        legend = c("Marriage-happiness"),
        ylim = c(0, 80))
```

## Gender and age
```{r,warning=FALSE,echo=FALSE,fig.cap="Figure 2: Happiness and age, men (blue) and women (red) "}
ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat == 4,], aes(x=factor(age), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2 & z$educat == 4,], aes(x=factor(age), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") + 
  scale_x_discrete("Age") +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,60)) +
  theme_bw()
```
*Sample restrictions: college educated men and women*

## Labour-market affiliation
```{r,fig.heigth=4,warning=FALSE,echo=FALSE,fig.cap="Figure 3: Happiness and labour-market affiliation, men (blue) and women (red)"}

z$workstatus <- NA
z$workstatus[z$wrkstat==1] <- "Full-time work"
z$workstatus[z$wrkstat==2] <- "Part-time work"
z$workstatus[z$wrkstat==7] <- "Keeping house"


ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") + 
  stat_summary(data = z[z$sex == 1 & z$educat ==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  scale_x_discrete("Work status", limits=c("Full-time work", "Part-time work", 
                "Keeping house")) +
  expand_limits(y=c(20,60)) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```
*Sample restrictions: college educated men and women*

## Family status and income
```{r,warning=FALSE,echo=FALSE,Cache=TRUE,fig.cap="Figure 4: Happiness and family constellation, men (blue) and women (red)"}

# Create career based on 50th cohort income percentile
z$career1 <- z$rinc > z$p50
z$career1 <- as.numeric(z$career1)
z$career1[is.na(z$career1)] <- 0 

# Create family constellation factors
z$meanhap <- NA
z$meanhap[z$family==0 & z$career1==0] <- "Low-inc., no family"
z$meanhap[z$family==0 & z$career1==1] <- "High-inc., no family"
z$meanhap[z$family==1 & z$career1==0] <- "Low-inc., family"
z$meanhap[z$family==1 & z$career1==1] <- "High-inc., family"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4 ,], aes(x=factor(meanhap), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(meanhap), y=vhappyb), 
             fun.y="mean", geom="point", col="Red") +
  scale_x_discrete("Family-work combination", limits=c("Low-inc., no family", 
              "High-inc., no family", "Low-inc., family", "High-inc., family")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```
*Sample restrictions: college educated men and women*

# Regression analysis

## Model 1: Interaction effects of marriage and job income {.smaller}
```{r, results='asis'}

M1a <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
         data = subset(z, sex==2 & educat == 4))
M2a <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) +as.factor(race) + as.factor(bdec), 
          data = subset(z, sex==1 & educat == 4))

stargazer::stargazer(M1a, M2a,
    type = 'html',
    omit = c("age", "agesq", "year", "bdec", "race"),
    omit.labels = c("Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men"),
    covariate.labels = c("High-income", "Married", "High-income*Married"),
    dep.var.labels   = "Very happy")
```    
*Sample restrictions: college educated men and women*

## Model 2: Double-click on married individuals {.smaller}
```{r, results='asis'}

# Create a variable for partner's income
z$othinccat <- trunc(z$def_othinc/10000)
z$othinccat[is.na(z$othinc)] <- -1

# Regressions
M3 <- lm(vhappyb ~ career1 + keepinghouse + age + agesq + as.factor(othinccat) + as.factor(year) + as.factor(race) +         as.factor(bdec), data = subset(z, sex==2 & educat == 4 & married == 1))

M4 <- lm(vhappyb ~ career1 + keepinghouse + age + agesq + as.factor(othinccat) + as.factor(year) + as.factor(race)          + as.factor(bdec), data = subset(z, sex==1 & educat == 4 & married == 1))

stargazer::stargazer(M3, M4,
    type = 'html',
    omit = c("othinccat", "age", "agesq", "year", "bdec", "race"),
    omit.labels = c("Partner's income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men"),
    covariate.labels = c("High-income", "Keeping house"),
    dep.var.labels   = "Very happy")
```    
*Sample restrictions: married, college educated men and women*


## Outlook

- Investigate drivers of when overall happiness and work satisfaction differ
  + Regression design where sample is limited to individuals reporting high job-satisfaction
  + Investigate effects of work intensity

