---
title: "Can we have it all?"
author: "Philip Unger & Philipp Staender"
date: "2 May 2016"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    number_sections: yes
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load packages and create BibTeX file for R-packages
PackagesUsed <- c("ggplot2", "repmis", "plyr", "dplyr", "MASS", "Hmisc", "interplot", "gridExtra", "repmis", "car", "reshape", "gmodels", "quantmod", "foreign")

# Set working directory
#setwd("~/Documents/CDA/collaborative_projects/Final_Project_P-P") # Unger
setwd("~/R_data/Final_Project_P-P") # Staender

# Load PackagesUsed
repmis::LoadandCite(PackagesUsed, file = "packages_pres.bib", install = FALSE)
```

```{r, include=FALSE}
if(!file.exists("data/data_final.rda")) {
  source('data_combine.R')
}

# Opens data_final.rda if not already loaded as an object
if(!exists("z.df" )) {
  load("data/data_final.rda")
}
```

## Intro | The debate - Can women have it all? 

*'I still strongly believe that women can “have it all” (and that men can too). ... But not today, not with the way America’s economy and society are currently structured.'*

<div class="black">
Anne-Marie Slaughter - Former director of policy planning for the U.S. State Department
</div>

*'Partner with the right person, because you cannot have a full career and a full life at home with the children if you are also doing all the housework and childcare.'*

<div class="black">
Sheryl Sandberg - Facebook's Chief Operating Officer
</div>


## Intro

<div class="black">
Research Question
</div>

*"Are career pursuits reconcilable with a happy life?"* 

<div class="black">
Hypotheses:
</div>

1. Women who pursue a career, and have success with respect to earnings relative to male peers, have equal life-satisfaction as women who chooses to establish a family.

2. Career women report job-satisfaction equal to their male peers, while reporting lower overall life happiness

3. Career women in earlier cohorts report lower life-satisfaction relative to later generations

## Research Design | Data

<div class="black">
General Social Survey
</div>
- cross-sectional survey of the adult population in the United States
- conducted from 1972 to 2014
- socio-economic, attitudinal and behavioural questions

<div class="black">
Current Population Survey
</div>
- cross-sectional representative survey


## Research Design | Methodology

<div class="black">
Methodology
</div>

## Figure 1

```{r, fig1,warning=FALSE,echo=FALSE,fig.cap="Distribution of reported happiness and job-satisfaction"}

happy_dist <- table(z$happy)/sum(table(z$happy))*100
satjob_dist <- table(z$satjob)/sum(table(z$happy))*100

par(mfrow=c(1,2)) 

barplot(happy_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness",
        ylab = "Frequency (percent)",
        legend = c("Happiness"),
        ylim = c(0, 80))  

barplot(satjob_dist, 
        col=c("navyblue"),
        xlab = "Reported job-satisfaction",
        legend = c("Job-satisfcation"),
        ylim = c(0, 80))
```

## Figure 5

```{r, fig5,fig.heigth=5,warning=FALSE,echo=FALSE,fig.cap="Happiness and labour-market affiliation (college educated men and women)"}

z$workstatus <- NA
z$workstatus[z$wrkstat==1] <- "Full-time work"
z$workstatus[z$wrkstat==2] <- "Part-time work"
z$workstatus[z$wrkstat==7] <- "Keeping house"


ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") + 
  stat_summary(data = z[z$sex == 1 & z$educat ==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  scale_x_discrete("Work status", limits=c("Full-time work", "Part-time work", 
                "Keeping house")) +
  expand_limits(y=c(20,60)) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```
- some comments
- some more commments

## Figure 7

```{r, fig6,warning=FALSE,echo=FALSE,fig.cap="Happiness and income level"}

z = z[z$year >= 1977 & z$year < 2012,]

z$income_status[z$career==1] <- "High-income"
z$income_status[z$career==0] <- "Low-income"

  # Following Bertrand (2013), respondents who report no income (Not applicable) is coded as
  # having low-income. Thus there are no N/A's in the career variables (see also below). 


z$career1 <- z$rinc > z$p50
z$career1 <- as.numeric(z$career1)
z$career1[is.na(z$career1)] <- 0 

z$income_status1[z$career1==1] <- "High-income"
z$income_status1[z$career1==0] <- "Low-income"

plot1 <- ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("High-income", "Low-income")) +
  scale_y_continuous("Percentage very happy") +
  ggtitle("A. Threshold: 25th percentile") +
  theme_bw()


plot2 <- ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status1), y=vhappyb),
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("High-income", "Low-income")) +
  scale_y_continuous("") +
  ggtitle("B. Threshold: 50th percentile") +
  theme_bw()
grid.arrange(plot1, plot2, ncol=2)
```

