---
title: "Untitled"
author: "Philipp Staender"
date: "8 May 2016"
output: html_document
---

```{r,results='asis',warning=FALSE,echo=FALSE}


z$sexcat <- NULL
z$sexcat[z$sex==1] <- "Male"
z$sexcat[z$sex==2] <- "Female"



z$spwrkcat <- NULL
z$spwrkcat[z$spwrksta==1] <- "Full-time work"
z$spwrkcat[z$spwrksta==2] <- "Part-time work"
z$spwrkcat[z$spwrksta==7] <- "Keeping house"

z$career1 <- z$rinc > z$p50
z$career1 <- as.numeric(z$career1)
z$career1[is.na(z$career1)] <- 0 

z$income_status1 <- NULL
z$income_status1[z$career1==1] <- "High-income"
z$income_status1[z$career1==0] <- "Low-income"

# Gender and Spouse Work Status
ge_sp_wrkst <- table(z$sexcat[z$educat==4], z$spwrkcat[z$educat==4])
knitr::kable(prop.table(ge_sp_wrkst, 1), caption = "Gender and Spouse Work Status",digits = 2)
```

```{r,results='asis',warning=FALSE,echo=FALSE}
# Gender, Income and Spouse Work Status
ge_sp_ca <- table(z$sexcat[z$educat==4 & z$married==1], z$income_status1[z$educat==4 & z$married==1], z$spwrkcat[z$educat==4 & z$married==1]) 
knitr::kable(ftable(prop.table(ge_sp_ca,c(1,2))), caption = "Gender and Spouse Work Status",digits = 2)
```

\begin{table}[]
\centering
\caption{Gender, income (p50) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 92.45     & 4.15      & 3.4           & 265  \\
       & Low-income  & 95.62     & 0.99      & 3.39          & 1713 \\
Male   & High-income & 42.28     & 33.56     & 24.16         & 745  \\
       & Low-income  & 54.86     & 25.06     & 20.08         & 1285 \\ \hline
\end{tabular}
\end{table}

\begin{table}[]
\centering
\caption{Gender, income (p25) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 94.1      & 3.04      & 2.86          & 559  \\
       & Low-income  & 95.63     & 0.78      & 3.59          & 1419 \\
Male   & High-income & 47.71     & 28.98     & 23.31         & 1201 \\
       & Low-income  & 53.92     & 27.02     & 19.06         & 829  \\ \hline
\end{tabular}
\end{table}

\begin{table}[]
\centering
\caption{Gender, income (p25) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{lllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 94.1      & 3.04      & 2.86          & 559  \\
       & Low-income  & 95.63     & 0.78      & 3.59          & 1419 \\
Male   & High-income & 47.71     & 28.98     & 23.31         & 1201 \\
       & Low-income  & 53.92     & 27.02     & 19.06         & 829  \\ \hline
\end{tabular}
\end{table}