---
title: "Can we have it all?"
author: "Philip Unger & Philipp Ständer"
date: "13 May 2016"
header-includes:
    - \usepackage{setspace}\doublespacing
output: 
  pdf_document:
    number_sections: yes
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    toc: yes
subtitle: How reconcilability of career pursuits and life satisfaction differs between women and men
bibliography: 
- main.bib
- packages.bib
---
\pagebreak

```{r, message=FALSE, include=FALSE}
# Load packages and create BibTeX file for R-packages
PackagesUsed <- c("ggplot2", "repmis", "plyr", "dplyr", "MASS", "Hmisc", "interplot", "gridExtra", "repmis", "car", "reshape", "gmodels", "quantmod", "foreign")

# Set working directory
 setwd("~/Documents/CDA/collaborative_projects/Final_Project_P-P") # Unger
# setwd("~/R_data/Final_Project_P-P") # Staender

# Load PackagesUsed and create .bib BibTeX file
repmis::LoadandCite(PackagesUsed, file = "packages.bib", install = FALSE)
```

```{r, include=FALSE}
if(!file.exists("data/data_sets/data_final.rda")) {
  source('data/data_combine.R')
}

# Opens data_final.rda if not already loaded as an object
if(!exists("z.df" )) {
  load("data/data_sets/data_final.rda")
}
```

# Descriptive results

## Inspecting the variables of interest

Two of the central variables for our research are reported happiness and job-satisfaction, which are based on the two questions: *"Taken all together, how would you say things are these days?"* and *"On the whole, how satisfied are you with the work you do?"*. Figure 1 shows the distribution of answers to the two questions. The former is measured on a three point scale (higher is better), of which around 60 % of the sample report a middle happiness level. The latter is measured on a four point scale, and Figure 1 shows that a majority of the sample report either moderate or high job satisfaction (score 3 and 4). 


```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of reported happiness and job-satisfaction"}

z$hapmarr <- NA
z$hapmarr[z$hapmar==3] <- 1
z$hapmarr[z$hapmar==2] <- 2
z$hapmarr[z$hapmar==1] <- 3

z$hapmar <- NULL
z <- rename(z, c(hapmarr="hapmar"))

happy_dist <- table(z$happy)/sum(table(z$happy))*100
satjob_dist <- table(z$satjob)/sum(table(z$happy))*100
hapmar_dist <- table(z$hapmar)/sum(table(z$happy))*100

par(mfrow=c(1,3)) 

barplot(happy_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness",
        ylab = "Frequency (percent)",
        legend = c("Happiness"),
        ylim = c(0, 80))  

barplot(satjob_dist, 
        col=c("navyblue"),
        xlab = "Reported job-satisfaction",
        legend = c("Job-satisfcation"),
        ylim = c(0, 80))

barplot(hapmar_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness with marriage",
        legend = c("Marriage-happiness"),
        ylim = c(0, 80))
```


### Average happiness in different survey years
The GSS is conducted between 1972 and 2014. Due to year specific events, unintended differences in the implementation of the survey or trends in overall happiness, there can be year-specific differences. Figure 2 shows the average share of the population who reports to be very happy (score of 3) across the survey years.

``` {r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Average reported happiness over survey year, 1972-2014"}

fig_survyear <- ddply(z, c("year"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_survyear <- na.omit(fig_survyear)

ggplot() +
  geom_point(data = fig_survyear, aes(x=year, y=vhappyb), color="Navyblue") +
  stat_smooth(data = fig_survyear, aes(x=year, y=vhappyb), method ="loess", col = "Navyblue") +
  scale_y_continuous("Percentage very happy") +
  scale_x_continuous("Survey Year") +
  expand_limits(y=c(25,45)) +
  theme_bw()
```

Figure 2 shows that there is considerable variation between years, and a weak negative trend particularly between 1972 and 1983. It is not directly possible to disentangle what can be attributed to random noise and what is caused by structural changes, however, it signifies that it is pragmatic to control for survey year in a regression design to control for survey year fixed effects. 

### Happiness and age
Figure 3 investigates the relationship between reported happiness and age for college educated men and women. In the GSS there is no apparent structural relationship between the share of respondents who report being very happy and age. Further, college educated women have a slightly higher average reported happiness level relative to men (38% vs. 34%).  

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and age (college educated men and women)"}

fig_age <- ddply(z[z$educat==4,], c("age", "sex"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_age <- na.omit(fig_age)

ggplot() +
  geom_point(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), color="Navyblue") + 
  geom_point(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), color="Red") +
  stat_smooth(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), method ="loess", se = FALSE, col="Navyblue") +
  stat_smooth(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), method ="loess", se = FALSE, col="Red") +
  expand_limits(y=c(20,60)) +
  scale_x_continuous("Age") +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,60)) +
  theme_bw()
```

## Respondent's happiness and income 

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of income"}

par(mfrow=c(1,2))
hist(z$def_rinc[z$sex == 1 & z$educat == 4],
     col = c("Navyblue"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated men)",
     main = ""
     )

hist(z$def_rinc[z$sex == 2 & z$educat ==4],
     col = c("Darkred"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated men)",
     main = ""
     )
```

**Needs to be populated**

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and respondent's income (college educated men and women)"}

par(mfrow=c(1,1))

ggplot() +
  stat_smooth(data = z[z$sex == 1 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), method ="loess", col="Navyblue") + 
  stat_smooth(data = z[z$sex == 2 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), method ="loess", col = "Red") +
  scale_x_continuous("Respondent's income") +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

**This section has to be populated** 

We should probably comment on:
* Women's happiness as home-going.
* That income is based on brackets, but deflated. 
* There are few observations with very high-income (high statistical dispersion). 

### Distribution of work-hours for full-time workers

Work intensity is another variable of interest for the final research design. "Hours worked last week" is available across the full survey from 1972-2014. This is an imperfect proxy for respondents average working hours over a full year, which unfortunately is not available. Figure 4 shows the frequency of hours worked last week for full-time workers. It is apparent that most respondents have worked around 40 hours a week, but also that there is considerable variance. 

```{r, fig.heigth=5,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Histogram of hours worked last week for full-time workers"}
par(mfrow=c(1,1)) 
satjob_dist <- table(z$satjob)/sum(table(z$happy))*100
hist(z$hrs1[z$working_ft==1],
     col=c("navyblue"),
     border = "White",
     xlab = "Hours worked last week (full-time employees)",
     main = ""
)   
```

## Work, household constallations & gender

In today's society the strive for gender equality often clashes with labour market realities and the persistence of traditional gender norms. Previous studies found that the reconcilability of having a family and pursuing a career is still difficult to achieve for women, while it is rather the norm for men [@hipp2015]. Thus, women who clash with traditional customs might face a “life happiness penalty”.

Figure 5 shows how reported happiness depends on labour-market affiliation for men (blue) and women (red) with a college degree. It shows that men are substantially more likely to report being very happy when in full-time employment relative to part-time employment, which is not the case for women. Further, both men and women report high happiness levels when keeping house. Note, in the full sample only 35 college educated men are keeping house (there are 650 women) and when looking at all men, the average share who reports being very happy while keeping house is only 24 %. 


```{r,fig.heigth=6,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and labour-market affiliation (college educated men and women)"}

z$workstatus <- NA
z$workstatus[z$wrkstat==1] <- "Full-time work"
z$workstatus[z$wrkstat==2] <- "Part-time work"
z$workstatus[z$wrkstat==7] <- "Keeping house"

ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="red") + 
  stat_summary(data = z[z$sex == 1 & z$educat ==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  scale_x_discrete("Work status", limits=c("Full-time work", "Part-time work", 
                "Keeping house")) +
  expand_limits(y=c(20,60)) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

Figure 6 shows the share of college educated men and women who report being very happy depending on whether they earn more than the 25th (panel A) or 50th (panel B) income percentile of college educated men in their age cohort. The graph suggests that women have the same propensity to be very happe regardless of whether they are high earners or not, whereas the difference is substantial for men. The difference for men is even more pronounced when  the threshold is set at the 50th percentile.

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and income level"}

z = z[z$year >= 1977 & z$year < 2012,]

z$income_status[z$career==1] <- "High-income"
z$income_status[z$career==0] <- "Low-income"

  # Following Bertrand (2013), respondents who report no income (Not applicable) is coded as
  # having low-income. Thus there are no N/A's in the career variables (see also below). 

z$career1 <- z$rinc > z$p50
z$career1 <- as.numeric(z$career1)
z$career1[is.na(z$career1)] <- 0 

z$income_status1 <- NULL
z$income_status1[z$career1==1] <- "High-income"
z$income_status1[z$career1==0] <- "Low-income"

plot1 <- ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("Low-income", "High-income")) +
  scale_y_continuous("Percentage very happy") +
  ggtitle("A. Threshold: 25th percentile") +
  theme_bw()


plot2 <- ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status1), y=vhappyb),
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("Low-income", "High-income")) +
  scale_y_continuous("") +
  ggtitle("B. Threshold: 50th percentile") +
  theme_bw()
grid.arrange(plot1, plot2, ncol=2)
```

\begin{table}[]
\centering
\caption{Gender, income (p50) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 92.45     & 4.15      & 3.4           & 265  \\
       & Low-income  & 95.62     & 0.99      & 3.39          & 1713 \\
Male   & High-income & 42.28     & 33.56     & 24.16         & 745  \\
       & Low-income  & 54.86     & 25.06     & 20.08         & 1285 \\ \hline
\end{tabular}
\end{table}

\begin{table}[]
\centering
\caption{Gender, income (p25) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 94.1      & 3.04      & 2.86          & 559  \\
       & Low-income  & 95.63     & 0.78      & 3.59          & 1419 \\
Male   & High-income & 47.71     & 28.98     & 23.31         & 1201 \\
       & Low-income  & 53.92     & 27.02     & 19.06         & 829  \\ \hline
\end{tabular}
\end{table}

### Marriage / family constellation, income and happiness

*<The below figure is presented in two versions based on marriage and family respectively. I think we should use the marriage, as this leads more directly into the first regression analysis.>*

Figure 7 differentiates between the four possible combinations of having a family (married and children) and having a high- or low-income job (defined as earning more than the 25th income percentile of college educated men in the respondent's cohort). Both college educated men and women report substantially higher happiness levels when having a family. When not having a family, higher income improves life satisfaction for both genders although the increase is larger for men. Gender differences become more pronounced when people have a family. With a family, women are happier when they are not in a high-income job, whereas the opposite is true for men. 

These descriptive results suggest that on average married couples with kids are  best-off when following a male bread-winner model, which conflicts with more progressive gender norms. Note, however, that the results could be driven by omitted factors such as assymetric total family income or age across the family constellations which also could affect subjective well-being. In our final analysis we seek to identify the factors that are driving these results.

### Two versions of similar graph: Family or marriage? 

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="V1: Happiness and marriage constellation (college educated subsample)"}
z$meanhap_m1 <- NA
z$meanhap_m1[z$family==0 & z$career1==0] <- "Low-inc., unmarried"
z$meanhap_m1[z$family==0 & z$career1==1] <- "High-inc., unmarried"
z$meanhap_m1[z$family==1 & z$career1==0] <- "Low-inc., married"
z$meanhap_m1[z$family==1 & z$career1==1] <- "High-inc., married"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4 ,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(meanhap_m1), y=vhappyb), 
             fun.y="mean", geom="point", col="Red") +
  scale_x_discrete("Family-work combination", limits=c("Low-inc., unmarried", 
              "High-inc., unmarried", "Low-inc., married", "High-inc., married")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="V2: Happiness and family constellation (college educated subsample)"}
z$meanhap <- NA
z$meanhap[z$family==0 & z$career==0] <- "Low-inc., no family"
z$meanhap[z$family==0 & z$career==1] <- "High-inc., no family"
z$meanhap[z$family==1 & z$career==0] <- "Low-inc., family"
z$meanhap[z$family==1 & z$career==1] <- "High-inc., family"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4 ,], aes(x=factor(meanhap), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(meanhap), y=vhappyb), 
             fun.y="mean", geom="point", col="Red") +
  scale_x_discrete("Family-work combination", limits=c("Low-inc., no family", 
              "High-inc., no family", "Low-inc., family", "High-inc., family")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```


# Regression Analysis

## Interaction effects of marriage and job income for working men and women

The correlations shown in Figure 5-8 are influenced by omitted factors. To control for some of the confounding factors that are observable, we replicate a linear probability model by Bertrand (2013) and estimate the effect of marriage and the interaction effect of marriage and having a high-paid job (career) on the binary variable of being very happy. While @bertrand2013 limits her analysis to college-educated women who are working, we also compare these findings to college educated men. The model controls for age, age-squared, the survey year, race and decade of birth.

Figure 9 shows the effect of marriage on the probability of being very happy for college educated men and women depending on job income. First, the effect of marriage is positive and significantly different from zero regardless of respondents' income level. The left panel shows that the effect of marriage on reported happiness is stronger for women who do not have a high-income job compared to women who do, as the interaction term between marriage and high job income is -0.07. Although this difference only is significant at the 10% level, job income seem to be much more important for the happiness of women compared to men, where having a high-paying job or not hardly influences the effect of marriage on happiness. 


```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Interaction effects of marriage and job income on life satisfaction"}

z$working_ft <- as.numeric(z$working_ft)
z$working_pt <- as.numeric(z$working_pt)

M1a <- lm(vhappyb ~ career*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
         data = subset(z, sex==2 & educat == 4))
M2a <- lm(vhappyb ~ career*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
          data = subset(z, sex==1 & educat == 4))

# compare men and women

interM1 <- interplot(M1a, var1 = "married", var2 = "career") +
  ggtitle("Working women") +
  xlab("1 = High-income") + 
  ylab("effect of marriage on life satisfaction") + 
  expand_limits(y=c(0.05,0.3)) +
  theme_bw()

interM2 <- interplot(M2a, var1 = "married", var2 = "career") +
  ggtitle("Working men") +
  xlab("1 = High-income") + 
  expand_limits(y=c(0.05,0.3)) +
  theme_bw()

grid.arrange(interM1, interM2, ncol = 2) 

```

## Cohorts and norms

**Populate with text.**

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and family constellation (college educated women)"}
z$meanhap1 <- NA
z$meanhap1[z$family==0 & z$career1==0] <- "Low-inc., no family"
z$meanhap1[z$family==0 & z$career1==1] <- "High-inc., no family"
z$meanhap1[z$family==1 & z$career1==0] <- "Low-inc., family"
z$meanhap1[z$family==1 & z$career1==1] <- "High-inc., family"

# Based on marriage rather than family

z$meanhap_m <- NA
z$meanhap_m[z$married==0 & z$career==0] <- "Low-inc., no family"
z$meanhap_m[z$married==0 & z$career==1] <- "High-inc., no family"
z$meanhap_m[z$married==1 & z$career==0] <- "Low-inc., family"
z$meanhap_m[z$married==1 & z$career==1] <- "High-inc., family"

z$meanhap_m1 <- NA
z$meanhap_m1[z$family==0 & z$career1==0] <- "Low-inc., no family"
z$meanhap_m1[z$family==0 & z$career1==1] <- "High-inc., no family"
z$meanhap_m1[z$family==1 & z$career1==0] <- "Low-inc., family"
z$meanhap_m1[z$family==1 & z$career1==1] <- "High-inc., family"

# - ggplot based on three cohorts
ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 4,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 5,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 6,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Purple") +
    scale_x_discrete("Family-work combination", limits=c("Low-inc., no family", 
                                                       "High-inc., no family", "Low-inc., family", "High-inc., family")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()

#- Take-aways: 1. With family and p25, we have a narrative that there are few generational changes. Issue, it cannot
#- be implemented in a regression design as the observations are way too few in the high-income, family group. 
#- 2. With p50, family, the point is different in high, fam, who "have it all". However, there is no significance. 
#- 3. Married and p25 the narrative is blurry. But we do seem to see a "bettering". 
```



# Discussion

# Conclusion

The preliminary analyses indicate that some variation in reported happiness is associated with job-affiliation and gender. Further, our descriptive results suggest that determinants for happiness, such as having a family or high job-income, differ in magnitude and direction between genders. This supports our initial assumption of differences in reconcilability of a career pursuit and a happy life. However, it remains a challenge to construct models which can attenuate problems of confounding factors. For the final project we intend to investigate in more detail how the intensity of work influences happiness and whether there is a trade-off between job satisfaction and overall hapiness.


# Software and packages used for the analysis
The analysis is done in R [@CiteR] with the use of the following packages: "ggplot2" [@R-ggplot2], "repmis" [@R-repmis], "plyr" [@R-plyr], "dplyr" [@R-dplyr], "MASS" [@R-MASS], "Hmisc" [@R-Hmisc], "interplot" [@R-interplot], "gridExtra" [@R-gridExtra], "car" [@R-car], "foreign" [@R-foreign], "gmodels" [@R-gmodels], "quantmod" [@R-quantmod] and "reshape" [@R-reshape].

# References