---
title: "Can we have it all?"
author: "Philip Unger & Philipp Ständer"
date: "13 May 2016"
header-includes:
    - \usepackage{setspace}\doublespacing
output: 
  pdf_document:
    number_sections: yes
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    toc: yes
    keep_tex: yes
subtitle: How reconcilability of career pursuits and life satisfaction differs between women and men
bibliography: 
- main.bib
- packages.bib
---
\pagebreak

```{r, message=FALSE, include=FALSE}
# Load packages and create BibTeX file for R-packages
PackagesUsed <- c("ggplot2", "repmis", "plyr", "dplyr", "MASS", "Hmisc", "interplot", "gridExtra", "repmis", "car", "reshape", "gmodels", "quantmod", "foreign", "stargazer")

# Set working directory
 setwd("~/Documents/CDA/collaborative_projects/Final_Project_P-P") # Unger
# setwd("~/R_data/Final_Project_P-P") # Staender

# Load PackagesUsed and create .bib BibTeX file
repmis::LoadandCite(PackagesUsed, file = "packages.bib", install = FALSE)
```

```{r, include=FALSE}
if(!file.exists("data/data_sets/data_final.rda")) {
  source('data/data_combine.R')
}

# Opens data_final.rda if not already loaded as an object
if(!exists("z.df" )) {
  load("data/data_sets/data_final.rda")
}
```

# Descriptive results

In this section we explore the variation in some of the main variables used in the analysis and illustrates correlations between reported happiness and survey-respondent characteristics that motivates the regression analyses. 

## Life-satisfaction measures in the GSS

Two of the central variables in the analysis are reported overall happiness and happiness of marriage, which are based on the two questions: *"Taken all together, how would you say things are these days?"* and *"Taking things all together, how would you desribe your marriage?"*. Figure 1 shows the distribution of answers to the two questions. Both are measured on a three point scale where higher is better. Panel A shows that approximately 40 % of the sample report high overall happiness, whereas panel B shows that a majority of the sample reports high marriage happiness. 

In the remainder of the paper we use binary variables to capture whether respondents report the highest happiness score or not on both measures. This eases interpretation, and as only a small portion of respondents report low-happiness on both measures, we do not disregard a substantial share of the variation in the variables. 


```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of reported happiness and job-satisfaction"}

z$hapmarr <- NA
z$hapmarr[z$hapmar==3] <- 1
z$hapmarr[z$hapmar==2] <- 2
z$hapmarr[z$hapmar==1] <- 3

z$hapmar <- NULL
z <- rename(z, c(hapmarr="hapmar"))

happy_dist <- table(z$happy)/sum(table(z$happy))*100
hapmar_dist <- table(z$hapmar)/sum(table(z$hapmar))*100

par(mfrow=c(1,2)) 

barplot(happy_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness",
        ylab = "Frequency (percent)",
        ylim = c(0, 80),
        main = "Panel A: Marriage happiness",
        font.main = 1)

barplot(hapmar_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness with marriage",
        ylim = c(0, 80),
        main = "Panel B: Marriage happiness",
        font.main=1)
```


## Reported happiness in different survey years
The GSS is conducted every or every other year between 1972 and 2014. Due to year specific events, unintended differences in the implementation of the survey or trends in overall happiness, there can be non-random, year-specific differences. Figure 2 shows the average share of the sample population who reports to be very happy across the survey years.

``` {r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Average reported happiness over survey year, 1972-2014"}

fig_survyear <- ddply(z, c("year"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_survyear <- na.omit(fig_survyear)

ggplot() +
  geom_point(data = fig_survyear, aes(x=year, y=vhappyb), color="Navyblue") +
  stat_smooth(data = fig_survyear, aes(x=year, y=vhappyb), method ="loess", col = "Navyblue") +
  scale_y_continuous("Percentage very happy") +
  scale_x_continuous("Survey Year") +
  expand_limits(y=c(25,45)) +
  theme_bw()
```

Figure 2 shows that there is considerable variation between years. Furthermore, there appears to be a negative trend between 1972 and 1985. It is not directly possible to disentangle what can be attributed to random noise and what is caused by structural changes, however, it signifies that it is pragmatic to control for survey year in the regression models. 

## Happiness and age
Figure 3 illustrates the relationship between reported happiness and age for college educated men and women. In the GSS there is no apparent structural relationship between the share of respondents who report being very happy and age. Furthermore, college educated women have a slightly higher average reported happiness level relative to men (38% vs. 34%).  

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and age (college educated men and women)"}

fig_age <- ddply(z[z$educat==4,], c("age", "sex"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_age <- na.omit(fig_age)

ggplot() +
  geom_point(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), color="Navyblue") + 
  geom_point(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), color="Red") +
  stat_smooth(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), method ="loess", se = FALSE, col="Navyblue") +
  stat_smooth(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), method ="loess", se = FALSE, col="Red") +
  expand_limits(y=c(20,60)) +
  scale_x_continuous("Age") +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,60)) +
  theme_bw()
```

## Respondent's income and reported happiness

Figure 4 shows the distribution of respondents' annual income in 2009 USD for college educated men and women. We can deduce that more men than women report high income. Furthermore, it shows that more women than men report no income at all, either due to unemployment or because they voluntarily keep house. 

```{r, fig.height=6, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of income"}

par(mfrow=c(2,1))
hist(z$def_rinc[z$sex == 1 & z$educat == 4],
     col = c("Navyblue"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated men)",
     main = ""
     )

hist(z$def_rinc[z$sex == 2 & z$educat ==4],
     col = c("Darkred"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated women)",
     main = ""
     )
```

Figure 5 shows the correlation between respondents income and the propensity to report high happiness for college educated women and men. For both genders there are a positive trend, which aligns with the majority of reserach on income and happiness [INSERT A REF]. A large share of women with very low or no income, however, also report high happiness. This reflects that married women who keeps house tend to report high happiness, which we will return to in the analysis. 

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and respondent's income (college educated men and women)"}

par(mfrow=c(1,1))

ggplot() +
  stat_smooth(data = z[z$sex == 1 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), method ="loess", col="Navyblue") + 
  stat_smooth(data = z[z$sex == 2 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), method ="loess", col = "Red") +
  scale_x_continuous("Respondent's income") +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

## Work, household constallations and gender

In this section we explore how reported happiness depend on labour market affiliation and family constellation for women and men, and discern gender differences. All figures are restricted to college educated women and men. 

Figure 6 shows how reported happiness depends on labour-market affiliation for men (blue) and women (red) with a college degree. It shows that men are substantially more likely to report being very happy when in full-time employment relative to part-time employment, which is not the case for women. Furthermore, both men and women report high happiness levels when keeping house. Note, however, that there are only 35 college educated men in the full sample who keep house, whereas there are 650 women. When looking at all men, the average share who reports being very happy while keeping house is only 24 %. 


```{r,fig.heigth=6,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and labour-market affiliation (college educated men and women)"}

z$workstatus <- NA
z$workstatus[z$wrkstat==1] <- "Full-time work"
z$workstatus[z$wrkstat==2] <- "Part-time work"
z$workstatus[z$wrkstat==7] <- "Keeping house"

ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="red") + 
  stat_summary(data = z[z$sex == 1 & z$educat ==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  scale_x_discrete("Work status", limits=c("Full-time work", "Part-time work", 
                "Keeping house")) +
  expand_limits(y=c(20,60)) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

A key variable in the research design is whether the respondents earn a high-income or not. To define high-income we use whether respondents earn more than the 50th income percentile of males in their education-age cohort. Figure 7 indicates that women have the same propensity to be very happe regardless of whether they are high earners or not, whereas the difference is substantial for men. [Footnote: The same is true if high-income is defined as the 25th income percentile for college educated men and women]. [Footnote: We only have data on cohort income from 1977 to 2010, and the sample is therefore restricted to these years]. 

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and income level, college educated men and women"}

z = z[z$year >= 1977 & z$year < 2012,]

  # Following Bertrand (2013), respondents who report no income (Not applicable) is coded as
  # having low-income. Thus there are no N/A's in the career variables (see also below). 

z$career1 <- z$rinc > z$p50
z$career1 <- as.numeric(z$career1)
z$career1[is.na(z$career1)] <- 0 

z$income_status1 <- NULL
z$income_status1[z$career1==1] <- "High-income"
z$income_status1[z$career1==0] <- "Low-income"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status1), y=vhappyb),
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("Low-income", "High-income")) +
  scale_y_continuous("") +
  theme_bw()
```

\begin{table}[]
\centering
\caption{Gender, income (p50) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 92.45     & 4.15      & 3.4           & 265  \\
       & Low-income  & 95.62     & 0.99      & 3.39          & 1713 \\
Male   & High-income & 42.28     & 33.56     & 24.16         & 745  \\
       & Low-income  & 54.86     & 25.06     & 20.08         & 1285 \\ \hline
\end{tabular}
\end{table}

\begin{table}[]
\centering
\caption{Gender, income (p25) and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 94.1      & 3.04      & 2.86          & 559  \\
       & Low-income  & 95.63     & 0.78      & 3.59          & 1419 \\
Male   & High-income & 47.71     & 28.98     & 23.31         & 1201 \\
       & Low-income  & 53.92     & 27.02     & 19.06         & 829  \\ \hline
\end{tabular}
\end{table}

# Analysis

<We need some sort of introduction to the section. Just a few sentences>- 

## Marriage constellation, income and happiness

Figure 8 illustrates how reported happiness depend on respondent's income and family affiliation. Specificially, we differentiate between the four combinations of marriage and having a high-income job or not. On average, both college educated men and women are more likely to report being very happy when they are married relative to not being married. Furthermore, when not married, both genders report higher happiness when having a high-income job. For married respondents, however, there are considerable gender differences. When married, men continue to report higher happiness when having high-income, whereas the opposite is true for women. This provides preliminary evidence on the thesis that women face difficulties in reconciling having a high-income job and marriage. 

**Figure: Happiness and marriage/income constellation.**

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and marriage constellation (college educated subsample)"}
z$meanhap_m1 <- NA
z$meanhap_m1[z$married==0 & z$career1==0] <- "Low-inc., unmarried"
z$meanhap_m1[z$married==0 & z$career1==1] <- "High-inc., unmarried"
z$meanhap_m1[z$married==1 & z$career1==0] <- "Low-inc., married"
z$meanhap_m1[z$married==1 & z$career1==1] <- "High-inc., married"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4 ,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(meanhap_m1), y=vhappyb), 
             fun.y="mean", geom="point", col="Red") +
  scale_x_discrete("Marriage-work combination", limits=c("Low-inc., unmarried", 
              "High-inc., unmarried", "Low-inc., married", "High-inc., married")) +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,50)) +
  theme_bw()
```


## Interaction effects of marriage and job income for working men and women

To investigate the gender differences illustrated in figure 8, we deploy different regression models to investigate the correlations further. Table x shows the effect of marriage, high-income and the interaction between the two on the binary variable of being very happy. Specification (3)-(4) include no control variables, and the reported coefficient are thus equivalent to results illustrated in figure 8. The regression design, however, allows us to deduce that the effect of marriage and having a high-income are significant for both genders at a 10 percent level or lower. Most interestingly, women report significantly lower happiness when both having high-income and are married, which is not the case for men. Women thus appear to face a substantial life-happiness penalty when "having it all". 

As explained in the research design section, a major methodological challenge is to create a valid counterfactual. Individuals in the different marriage-work constellations are likely to differ from each other in multiple dimensions that correlate with reported happiness. In specification (1) and (2) we control for potential confounding factors that are observable: age, age-squared, survey year, race and cohort (decade of birth). It is apparent that the results are robust to their inclusion. 

**Table 1: Happiness, marriage and high-income**

```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

T1.1 <- lm(vhappyb ~ career1*married, 
         data = subset(z, sex==2 & educat == 4))
T1.2 <- lm(vhappyb ~ career1*married, 
          data = subset(z, sex==1 & educat == 4))
T1.3 <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
         data = subset(z, sex==2 & educat == 4))
T1.4 <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
          data = subset(z, sex==1 & educat == 4))


stargazer(T1.3, T1.4, T1.1, T1.2,
    type = 'text',
    header = FALSE,
    title = "Happiness, marriage and income for college educated men and women",
    omit.stat = c("f", "rsq", "ser"),
    omit = c("age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men"),
    digits = 2,
    covariate.labels = c("High-income", "Married", "High-income*Married"),
    dep.var.labels   = "Very happy",
    notes = "Models are restricted to college educated men and women")
```    


## Double-click on married respondent's: How does high-income differ between genders? 

A considerable source of sample heterogeneity is that married and non-married individuals could differ in characteristics that are not observable and that affects reported happiness. To address this we circumscribe the sample to married individuals in table Y. Furthermore, we distinguish between three types of individuals: repsondent's with high-income, respondent's with low-income and respondens who keeps house. The latter category only includes few college educated, married men (xx% of the sample), but a substantial amount of college educated, married women (xx%). 

Coloumn (1)-(2) further includes for the survey respondent's partner's income, as this is likely to affect material well-being and thus reported happiness as well as the respondent's labour market affiliation, as well as the controls used in table X. [footnote: Partner's income is controlled for with dummy variables for $10.000 income brackets]. 

From specification (2) we observe that married, college educated men report significantly higher happiness when earning high-income relative to low-income, and that men who keep house on average reports the lowest happiness, though it is not significantly different from zero. Women, on the other hand, do not report significantly higher happiness when being both married and having a high-income job, and though the coefficient is positive, it is of relatively small magnitude. Women who keep house, however, report significantly higher happiness than women with low-income at the 10 percent level. One explanation for this could be that some women who work part-time would prefer to keep-house, but are not able to for financial reasons. This should be mitigated by controlling for the spouses income, but this does not include a measure of family wealth. 

Table Y thus paints a similar picture to table X, which is that men are the happiest when they are both marrid and have a high-income job, whereas it is not the case for women. While there is no direct penalty for having a high-income relative to a low-income job, reported happiness is the highest when keeping house. 

**Table 2: Double-click on married individuals**
```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

# Create a variable for partner's income
z$othinccat <- trunc(z$def_othinc/10000)
z$othinccat[is.na(z$othinc)] <- -1

# Regressions
T2.1 <- lm(vhappyb ~ career1 + keepinghouse + kid,
           data = subset(z, sex==2 & educat == 4 & married == 1))

T2.2 <- lm(vhappyb ~ career1 + keepinghouse + kid, 
           data = subset(z, sex==1 & educat == 4 & married == 1))

T2.3 <- lm(vhappyb ~ career1 + keepinghouse + kid + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==2 & educat == 4 & married == 1))

T2.4 <- lm(vhappyb ~ career1 + keepinghouse + kid + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==1 & educat == 4 & married == 1))

stargazer(T2.3, T2.4, T2.1, T2.2,
    type = 'text',
    header = FALSE,
    title = "Happiness and work-status for married, college educated women and men",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("othinccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Partner's income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men"),
    covariate.labels = c("High-income", "Keeping house", "Child"),
    dep.var.labels   = "Very happy",
    notes = "Models include married, college educated men and women"
   )
```  

## Effect on happiness of young children

**Table 3: Happiness and young children** 
```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

# Variables
z$young_child <- 0
z$young_child[z$old1 <= 3 | z$old2 <= 3 | z$old3 <= 3 | z$old4 <= 3 | z$old5 <= 3 | z$old6 <= 3 
              | z$old7 <= 3 | z$old8 <= 3 | z$old9 <= 3 | z$old10 <= 3] <- 1

# Regressions
T3.1 <- lm(vhappyb ~ career1 + keepinghouse,
           data = subset(z, sex==2 & educat == 4 & family == 1))

T3.2 <- lm(vhappyb ~ career1 + keepinghouse, 
           data = subset(z, sex==1 & educat == 4 & family == 1))

T3.3 <- lm(vhappyb ~ career1 + keepinghouse + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==2 & educat == 4 & family == 1))

T3.4 <- lm(vhappyb ~ career1 + keepinghouse + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==1 & educat == 4 & family == 1))

T3.5 <- lm(vhappyb ~  keepinghouse*young_child + young_child*career1 + as.factor(othinccat)  + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex==2 & educat == 4 & family == 1))

T3.6 <- lm(vhappyb ~ young_child*career1 + as.factor(othinccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
             data = subset(z, sex==1 & educat == 4 & family == 1))

stargazer(T3.1, T3.2, T3.3, T3.4, T3.5, T3.6,
    type = 'latex',
    header = FALSE,
    title = "Happiness and young children, college educated women and men with a family",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("othinccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Partner's income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men", "Women", "Men"),
    covariate.labels = c("High-income", "Keeping house", "Young child", "Keeping House*Young child", "High-income*Young child"),
    dep.var.labels   = "Very happy",
    notes = "Models are restricted to married + child, college educated men and women")
```    

**Needs to be populated**

There story remains relatively weak. We don't have a significant negative interaction term, partly because we have less than 50 individuals that have both career1 and a young child.

Nonetheless, we can argue that at least it seems to be the career individuals with a young child that reports low  life hapiness. 

Do we have a problem of maternity leave? I guess they are excluded as they wont earn much. 

## Satisfaction with marriage and family constellation. 

**Table 4: Marriage happiness and spouse working conditions**

```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}
# Variables

z$inccat <- trunc(z$def_inc/10000)
z$inccat[is.na(z$def_inc)] <- -1

z$spouse_ft <- NULL
z$spouse_ft[z$spwrksta==1] <- 1
z$spouse_ft[z$spwrksta==2 | z$spwrksta==7] <- 0

z$spouse_pt <- NULL
z$spouse_pt[z$spwrksta==2] <- 1
z$spouse_pt[z$spwrksta==1 | z$spwrksta==7] <- 0

z$spouse_home <- NULL
z$spouse_home[z$spwrksta==7] <- 1
z$spouse_home[z$spwrksta==1 | z$spwrksta==2] <- 0

##Marriage happiness
z$vhapmarb <- NA
z$vhapmarb[z$hapmar==1 | z$hapmar==2] <- 0
z$vhapmarb[z$hapmar==3] <- 100

# Regressions

# T4.1 <- lm(vhappyb ~ spouse_ft + kid ,  data = subset(z, sex == 2 & educat == 4 & married ==1))

# T4.2 <- lm(vhappyb ~ spouse_ft + spouse_home + kid, data = subset(z, sex == 1 & educat == 4 & married ==1))

# T4.3 <- lm(vhappyb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec),  data = subset(z, sex == 2 & educat == 4 & married ==1))

# T4.4 <- lm(vhappyb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), data = subset(z, sex == 1 & educat == 4 & married ==1))

# Marriage happiness

T5.1 <- lm(vhapmarb ~ spouse_ft + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex == 2 & educat == 4 & married ==1))

T5.2 <- lm(vhapmarb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex == 1 & educat == 4 & married ==1))


T5.3 <- lm(vhapmarb ~ spouse_ft + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 2 & educat == 4 & married ==1 & career1 == 1))

T5.4 <- lm(vhapmarb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 1 & educat == 4 & married ==1 & career1 == 1))

T5.5 <- lm(vhapmarb ~ spouse_ft + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 2 & educat == 4 & family ==1 & career1 == 1))

T5.6 <- lm(vhapmarb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 1 & educat == 4 & family ==1 & career1 == 1))

stargazer(T5.1, T5.2, T5.3, T5.4, T5.5, T5.6,
    type = "latex",
    header = FALSE,
    title = "Marriage happiness and spouse's work-status",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("inccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Family income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men", "Women", "Men"),
    covariate.labels = c("Spouse FT", "Spouse Home", "Children"),
    dep.var.labels   = "Very happy (marriage)",
    notes = "Models are restricted to married, college educated men and women")
```
**We need to shine up the table and indicate (3) and (4) is for career1 == 1**

Text: The story needs to be pushed here again. Problem is that people presumably choose the career constellations they believe they will be happy in. 

We should probably add the contingency tables here. 

Question to be answered: Do we also report life happiness and family constellation? Problem is that career women actually prefer their husband to not work full-time :/ It could work as a story, but does hardly align with the lower marriage happiness. 

Another meth. problem: A full-time job is not necessarily intensive. 40 hours a week is not unmanageble. Problem is that we don't have a better variable. 

## Cohorts and norms

**Populate with text.**

**Figure: Cohortian differences**

```{r,fig.height=6, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and family constellation (college educated women)"}
z$meanhap1 <- NA
z$meanhap1[z$family==0 & z$career1==0] <- "Low-inc., no family"
z$meanhap1[z$family==0 & z$career1==1] <- "High-inc., no family"
z$meanhap1[z$family==1 & z$career1==0] <- "Low-inc., family"
z$meanhap1[z$family==1 & z$career1==1] <- "High-inc., family"

# Marriage rather than family
z$meanhap_m1 <- NA
z$meanhap_m1[z$married==0 & z$career1==0] <- "Low-inc., unmarried"
z$meanhap_m1[z$married==0 & z$career1==1] <- "High-inc., unmarried"
z$meanhap_m1[z$married==1 & z$career1==0] <- "Low-inc., married"
z$meanhap_m1[z$married==1 & z$career1==1] <- "High-inc., married"

# - ggplot based on three cohorts
genplot1 <- ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 4,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 5,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 6,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Purple") +
    scale_x_discrete("Family-work combination", limits=c("Low-inc., unmarried", "High-inc., unmarried", 
                                                         "Low-inc., married", "High-inc., married")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()

genplot2 <- ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 4,], aes(x=factor(meanhap1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 5,], aes(x=factor(meanhap1), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 6,], aes(x=factor(meanhap1), y=vhappyb), 
               fun.y="mean", geom="point", col="Purple") +
    scale_x_discrete("Family-work combination", limits=c("Low-inc., no family", 
                                                       "High-inc., no family", "Low-inc., family", "High-inc., family")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()

grid.arrange(genplot1, genplot2, ncol = 1) 
```



# Discussion

# Conclusion

The preliminary analyses indicate that some variation in reported happiness is associated with job-affiliation and gender. Further, our descriptive results suggest that determinants for happiness, such as having a family or high job-income, differ in magnitude and direction between genders. This supports our initial assumption of differences in reconcilability of a career pursuit and a happy life. However, it remains a challenge to construct models which can attenuate problems of confounding factors. For the final project we intend to investigate in more detail how the intensity of work influences happiness and whether there is a trade-off between job satisfaction and overall hapiness.


# Software and packages used for the analysis
The analysis is done in R [@CiteR] with the use of the following packages: "ggplot2" [@R-ggplot2], "repmis" [@R-repmis], "plyr" [@R-plyr], "dplyr" [@R-dplyr], "MASS" [@R-MASS], "Hmisc" [@R-Hmisc], "interplot" [@R-interplot], "gridExtra" [@R-gridExtra], "car" [@R-car], "foreign" [@R-foreign], "gmodels" [@R-gmodels], "quantmod" [@R-quantmod] and "reshape" [@R-reshape].

# References