---
title: "Can we have it all?"
author: "Philip Unger & Philipp Ständer"
date: "13 May 2016"
header-includes:
    - \usepackage{setspace}\doublespacing
output: 
  pdf_document:
    number_sections: yes
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    toc: yes
subtitle: Reconcilability of career pursuits and life satisfaction for women and men
bibliography: 
- main_paper.bib
- packages_paper.bib
---
\pagebreak

```{r, message=FALSE, include=FALSE}
# Load packages and create BibTeX file for R-packages
PackagesUsed <- c("ggplot2", "repmis", "plyr", "dplyr", "MASS", "Hmisc", "interplot", "gridExtra", "repmis", "car", "reshape", "gmodels", "quantmod", "foreign", "stargazer")

# Set working directory
try(setwd("~/Documents/CDA/collaborative_projects/Final_Project_P-P"), silent = TRUE)
try(setwd("~/R_data/Final_Project_P-P"), silent = TRUE)

# Load PackagesUsed and create .bib BibTeX file
repmis::LoadandCite(PackagesUsed, file = "packages_paper.bib", install = FALSE)
```

```{r, include=FALSE}
if(!file.exists("data/data_sets/data_final.rda")) {
  source('data/data_combine.R')
}

# Opens data_final.rda if not already loaded as an object
if(!exists("z.df" )) {
  load("data/data_sets/data_final.rda")
}
```

# Introduction 

*‘Can we have it all?’* The question of reconciling family and career has been at the heart of the gender equality debate in the United States (US) and other advanced economies. In the US, there are today more women than men who graduate from college and it seems natural to assume that most young women that make a high financial investment in education also seek to pursue a career [@goldin2006]. This is in line with contemporary research on life goals. Allmendinger (forthcoming), for instance, found based on a German survey that employment was the factor that most respondent, both males and females, regarded as ‘highly important’, more so than any other goal or value in life [^zeit].

[^zeit]: Preliminary results were published by @zeit2016.

Anne-Marie Slaughter [-@slaughter2012], former Director of Policy Planning for the U.S. State Department, wrote in an influential essay: *“I still strongly believe that women can ‘have it all’ (and that men can too). ... But not today, not with the way America’s economy and society are currently structured.”* This impression is in line with empirical evidence that indicates that particularly women find it difficult to reconcile work and family life, especially with respect to care responsibilities. Across industrialized countries, a majority of partnerships still end up in a single-breadwinner constellation (predominantly male) with asymmetric division of wage and care work [@hipp2015]. Furthermore, research has shown that an unequal distribution of paid- and care-work among partners is also matched by self-reported life-satisfaction, meaning that women report higher life-satisfaction when working part-time, while men prefer to work full-time [@booth2008]. @bertrand2013 further shows that college educated women with a high-income job report lower life-satisfaction on average than women with similar characteristics who work part time or not at all. 

These observations are in conflict with the strive of highly educated women to combine both career and family. In this paper we wish to investigate on survey data from the US whether reconcilability of career and family differs between college educated women and men, and explore some of the potential drivers. Consequently, we address the research question: *"Are career pursuits reconcilable with a happy life?"* We approach the research question in four steps. First, we show across different specifications and samples that women report lower life-satisfaction when having both a high-income job and a family, whereas this is not the case for men. Second, we show that the reconciliation of a high-income job and a family is particularly difficult for women who have a young child, and that it is not the case for men. Third, we investigate how marital happiness depends on the spouse's work status. We find that women tend to report higher happiness when their partner work full-time, while high-earning men report higher marital happiness their partners to stay at home or work part-time. Last, we examine whether younger female cohorts are more able to reconcile work and family than older female cohorts. Our descriptive results indicate that this indeed is the case, which could suggest a change in gender norms.

The paper proceeds as follows: In section two we review the relevant literature on gender-specific determinants of life-satisfaction. In the third section we present our empirical strategy and discuss methodological challenges. The fourth section presents descriptive results on key variables. In the fifth section we present our analysis. The sixth section concludes.

# Literature review

The following literature review provides a framework to analyse the gender-specific links between family constellations, job-affiliations and life-satisfaction. We first introduce economic theory of gender specialization as well as gender identity theory, second, we review findings of the literature about gender-specific work status preferences and third, we present existing evidence on the impact of family constellations on life-satisfaction.

Why do men and women distribute care and paid work unevenly instead of sharing both responsibilities? Economic theory regards the division of wage and household work as a form of economic specialization [@becker1965; -@becker1985] and an outcome of a bargaining process [@blood1995; @lundberg1996]. From a purely economic perspective, the role specialization among couples is based on their relative human capital and related income prospects. However, a comparative advantage perspective can only explain a fraction of the real division of wage and care work between men and women. 

The gender identity hypothesis proposed by @akerlof2000 assumes that utility functions are determined by “a person’s sense of self” and therefore differ according to gendered social norms and expectations. A study by @booth2008 lends support to the identity hypothesis by finding that British women reach highest job satisfaction and satisfaction with working-hours in part-time rather than full-time jobs, while men prefer to work full-time. Surprisingly, however, women’s life satisfaction is hardly affected by the hours they work. These results pose a puzzle: A preference for part-time employment with limited career perspectives seems counter-intuitive given the increasing investment of women into acquiring higher education.

@bertrand2013 investigates the relationship between career, family and well-being for college-educated women. The study finds that having a family contributes more strongly to women’s happiness than having a career. Interestingly, the study also finds that reported happiness is higher for women who have a family and no career relative to women with both. One possible explanation is Hochschild’s [-@hochschild1990] ‘second shift’ finding. His study shows that the main responsibility for housework rests disproportionately on women regardless of whether the household has a male or female breadwinner. 

The irreconcilability thesis is supported empirical evidence on an unequal distribution of care and paid work. @paull2008 finds that many women shift from full-time to part-time employment after the first child arrived, which significantly reduces the ability to pursue a career. @hipp2015 shows that differences in working time within couples are still significant across industrialized countries, and that working time differences are even more pronounced when couples are married and have children. However, when women work in an occupation with similar or higher status compared to their partner's occupation the working time differences decrease.

The literature hints to an interplay of gender norms and societal structures such as workplace arrangements that make it harder for women to reconcile a career with having a family. Nonetheless, the literature struggles to identify the drivers of a potential life-happiness penalty for women more precisely. Those studies that looked specifically at highly educated women show that women with higher education seem more likely to attempt to reconcile career and family [@hipp2015], but nonetheless struggle to derive higher life-satisfaction than women who do not attempt to pursue a career [@bertrand2013].

# Research Design

## Data 

The quantitative analyses use two primary data sources: the General Social Survey (GSS) and the Current Population Survey (CPS). 

* The GSS is a representative cross-sectional survey of the adult population in the United States, which has been conducted annually or biennially between 1972 and 2014. The survey tracks a wide-range of socio-economic, attitudinal and behavioural questions. The most important aspects for this research is reported life-satisfaction in different dimensions (overall happiness, marriage happiness and job-satisfaction), which can be linked to demographic, work and family characteristics.[CPE_foot] 

* The CPS is a nationwide cross-sectional representative survey with about 60.000 monthly respondents, and it is one of the primary sources of labour force statistics in the US. We use data from the CPS on income percentiles in every year for college educated men in each five-year age group (25-29, 30-34,...). The data is collected from the replication file of @bertrand2013.[^CPS_foot] 

[^CPS_foot]: Bertrand only includes the income percentiles between 1977-2010, and all analyses using data from the CPS are thus limited to these years. 

[^CPE_foot]: We use the Personal Consumption Expenditure chained index from the Federal Reserve Bank of St. Louis to adjust nominal income variables for inflation. 

## Empirical strategy and methodological challenges

In our empirical analyses, we will rely on graphical depictions as well as a relatively simple regression design. In all regression specifications the dependent variable will be binary, and following “Mostly Harmless Econometrics” [@angrist2008] we use a linear regression model rather than a logistic or a probit model. The deployed regressions are thus equivalent to linear probability models. Furthermore, in all specifications we pool survey respondents across all survey-years to maximize the available number of observations.  

Below we outline the major methodological caveats in our research design, which limits a causal interpretation of our results. 

**Subjective well-being data and statistical inference**  

To analyse how work affiliation and family constellations affect life-satisfaction, we rely on direct reports on subjective well-being (SWB) from the GSS. Using SWB data presents some methodological concerns, as insights from behavioural economics question the robustness of SWB data [@kahneman2006; @bertrand2013]. @kahneman2006, for instance, proposes that the effect of substantial life changes on subjective well-being are temporary. One explanation is that individuals gradually adjust their well-being aspirations; even when individuals achieve life-goals. Another issue relates to individuals ability to gauge their well-being. @nikolova2015, for instance, found that when survey participants were asked two similar life-satisfaction questions at different points in time during the same survey, 14 per cent of the individuals reported significantly different levels of well-being. The methodological caveats relating to SWB-measures should not be overlooked, but SWB-data nonetheless present an opportunity to explore correlations between individual characteristics and life-satisfaction directly. Furthermore, there is no consensus that any other measure is superior. 

**Operationalisation of career-pursuits**  

To analyze the relationship between career pursuits and life-satisfaction, we face a challenge in operationalizing when an individual pursues a career. In related academic literature, career pursuits have been proxied by income thresholds [@goldin2004; @bertrand2013]. Bertrand (2013), for instance, uses whether individuals earn more than the 25th income percentile of males in their age-education cohort. These measures, however, neglect that individuals can have career ambitions in low-income jobs, and that not all high earners are pursuing a career. Unfortunately, the GSS does not include variables that allow for a more nuanced operationalization of career pursuits.[^whs_foot] In this paper we use a similar approach to Bertrand (2013), specifically whether the survey respondent earns more than 50 percent of the income of males in their age-education cohort. Earning a high-income is likely to be a decent proxy for career pursuits, but we interpret it as merely indicating whether the survey-respondent has a high-income job or not. 

[^whs_foot]: The only measure for work-intensity included in the GSS is hours worked last week, which is an imperfect proxy for the average amount of hours worked over an entire year.  

**Sample heterogeneity**  
In investigating the relationship between family constellations, career pursuits and reported well-being measures, a major methodological limitation arises in addressing unobservable heterogeneity across different groups of individuals. For instance, when comparing reported life-satisfaction between women pursuing a career and women who do not, we ultimately face the problem that the two groups of individuals will differ in other characteristics that affect life-satisfaction than just career pursits. Hence, the presence of unobservable heterogeneity and the lack of a panel structure make it hard to construct a valid life-satisfaction counterfactual. Furthermore, one would expect that individuals at least to some extent choose the family and work-constellations which they believe will provide them the highest life-satisfaction. Thus, even if married women who pursue a career report lower life-satisfaction on average than married women who stay at home, it would be a fallacy to argue that the career-women would be happier if they quit their job. 

To minimise the issue of heterogeneity, we follow the approach of @bertrand2013 and limit our analysis to college educated survey respondents, as people with a college education are more likely to pursue a career and are likely to substantially differ from non-college educated individuals. Furthermore, we step-wise limit our sample size to specific sub-groups, for instance married individuals, individuals who are married and have children, and individuals with a high-income job.[^het_foot] 

[^het_foot]: Even though limiting the sample to specific groups attenuates the issue of sample heterogeneity, it introduces a problem of small sample sizes. When comparing individuals with specific educational attainment, labour-force participation, etc., the sample size of the GSS can be reduced to a few hundred observations.

**Interpretation of result**  
Due to the methodological challenges, the scope of the paper is not to establish a causal relationship between family constellations, career pursuits and well-being. It should be interpreted as an attempt to identify correlations that can help us understand why reconciling career and family responsibilities are problematic. Furthermore, by comparing women to men, we are allowed to get a better understanding of gender differences that can help explain why women appear to struggle more from reconciling career pursuits with family life.

# Descriptive results

In this section we explore the variation in some of the main variables used in the analysis. Furthermore, we illustrate correlations between reported happiness and survey-respondent characteristics, which we use to motivate the regression analyses. 

## Life-satisfaction measures in the GSS

Two of the central variables in the analysis are reported overall happiness and happiness of marriage, which are based on the two questions: *"Taken all together, how would you say things are these days?"* and *"Taking things all together, how would you describe your marriage?"*. Figure 1 shows the distribution of answers to the two questions. Both are measured on a three point scale where higher is better. Panel A shows that approximately 40 % of the sample report high overall happiness, whereas panel B shows that a majority of respondents report high marital happiness. 

In the remainder of the paper we use binary variables to capture whether respondents report the highest happiness score or not on both measures. This eases interpretation, and as only few respondents report low-happiness on both measures, we do not disregard a substantial share of the variation in the variables.


```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of reported happiness and job-satisfaction"}

happy_dist <- table(z$happy)/sum(table(z$happy))*100
hapmar_dist <- table(z$hapmar)/sum(table(z$hapmar))*100

par(mfrow=c(1,2)) 

barplot(happy_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness",
        ylab = "Frequency (percent)",
        ylim = c(0, 80),
        main = "Panel A: Marriage happiness",
        font.main = 1)

barplot(hapmar_dist, 
        col=c("navyblue"),
        xlab = "Reported happiness with marriage",
        ylim = c(0, 80),
        main = "Panel B: Marriage happiness",
        font.main=1)
```


## Reported happiness in different survey years
The GSS is conducted every or every other year between 1972 and 2014. Due to year specific events, unintended differences in the implementation of the survey or trends in overall happiness, there can be non-random, year-specific differences. Figure 2 shows that there is considerable variation in the share of individuals who report high-happiness across survey-years. Furthermore, there appears to be a negative trend between 1972 and 1985. It is not directly possible to disentangle what can be attributed to random noise and what is caused by structural changes, however, it signifies that it is pragmatic to control for survey year in the regression models. 

``` {r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and survey year, 1972-2014"}

fig_survyear <- ddply(z, c("year"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_survyear <- na.omit(fig_survyear)

ggplot() +
  geom_point(data = fig_survyear, aes(x=year, y=vhappyb), color="Navyblue") +
  stat_smooth(data = fig_survyear, aes(x=year, y=vhappyb), method ="loess", col = "Navyblue") +
  scale_y_continuous("Percentage very happy") +
  scale_x_continuous("Survey Year") +
  expand_limits(y=c(25,45)) +
  theme_bw()
```

## Happiness and age
Figure 3 illustrates the relationship between reported happiness and age for college educated men and women. We can observe that college educated women have a slightly higher propensity to report being very happy relative to men (38% vs. 34%), but there is no apparent structural relationship between age and happiness in the GSS. 

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and age (college educated men and women)"}

fig_age <- ddply(z[z$educat==4,], c("age", "sex"), summarise,
              vhappyb = mean(vhappyb, na.rm=TRUE))

fig_age <- na.omit(fig_age)

ggplot() +
  geom_point(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), color="Navyblue") + 
  geom_point(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), color="Red") +
  stat_smooth(data = fig_age[fig_age$sex == 1,], aes(x=age, y=vhappyb), method ="loess", 
              se = FALSE, col="Navyblue") +
  stat_smooth(data = fig_age[fig_age$sex == 2,], aes(x=age, y=vhappyb), method ="loess", 
              se = FALSE, col="Red") +
  expand_limits(y=c(20,60)) +
  scale_x_continuous("Age") +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,60)) +
  theme_bw()
```

## Respondent's income and reported happiness

Figure 4 shows the distribution of respondents' annual income in 2009 USD for college educated men and women. We can deduce that more men than women have a high income. Furthermore, it shows that more women than men report no income at all, either due to unemployment or because they voluntarily keep house. 

```{r, fig.height=6, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Distribution of income"}

par(mfrow=c(2,1))
hist(z$def_rinc[z$sex == 1 & z$educat == 4],
     col = c("Navyblue"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated men)",
     main = ""
     )

hist(z$def_rinc[z$sex == 2 & z$educat ==4],
     col = c("Darkred"),
     border = "White",
     xlim = c(0,250000),
     breaks = 20,
     xlab = "Respondent's income (College educated women)",
     main = ""
     )
```

Figure 5 plots the correlation between respondents income and the propensity to report high happiness for college educated women and men. For both genders there is a positive trend, which aligns with the majority of research on income and subjective well-being [@stevenson2013]. Furthermore, a large share of women with very low or no income also reports high happiness. This reflects that women who ‘keep house’ tend to report high happiness, which we will return to in the analysis. 

```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and respondent's income (college educated men and women)"}

par(mfrow=c(1,1))

ggplot() +
  stat_smooth(data = z[z$sex == 1 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), 
              method ="loess", col="Navyblue") + 
  stat_smooth(data = z[z$sex == 2 & z$educat ==4,], aes(x=def_rinc, y=vhappyb), 
              method ="loess", col = "Red") +
  scale_x_continuous("Respondent's income") +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

## Work, household constellations and gender

In this section we explore how reported happiness depend on labour market affiliation and family constellation for women and men, and discern gender differences. All figures are restricted to college educated women and men. 

Figure 6 shows how reported happiness depends on labour-market affiliation for men (blue) and women (red) with a college degree. It shows that men are substantially more likely to report being very happy when in full-time employment relative to part-time employment, which is not the case for women. Furthermore, both men and women report high happiness levels when keeping house. Note, however, that there are only 35 college educated men in the full GSS sample who keep house, whereas there are 650 women. When looking at all men, the average share who reports being very happy while keeping house is only 24 %.

```{r,fig.heigth=6,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and labour-market affiliation (college educated men and women)"}

ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="red") + 
  stat_summary(data = z[z$sex == 1 & z$educat ==4,], aes(x=factor(workstatus), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  scale_x_discrete("Work status", limits=c("Full-time work", "Part-time work", 
                "Keeping house")) +
  expand_limits(y=c(20,60)) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

A key variable in the analysis is whether the respondents earn a high-income or not. To define high-income we use whether respondents earn more than the 50th income percentile of males in their education-age cohort. Figure 7 indicates that women have the same propensity to be very happy regardless of whether they are high earners or not, whereas the difference is substantial for men. 

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and income level (college educated men and women)"}

z = z[z$year >= 1977 & z$year < 2012,]

z$income_status1 <- NULL
z$income_status1[z$career1==1] <- "High-income"
z$income_status1[z$career1==0] <- "Low-income"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4,], aes(x=factor(income_status1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") + 
  stat_summary(data = z[z$sex == 2  & z$educat==4,], aes(x=factor(income_status1), y=vhappyb),
               fun.y="mean", geom="point", col="Red") + 
  expand_limits(y=c(25,45)) +
  scale_x_discrete("High- or low-income job", limits=c("Low-income", "High-income")) +
  scale_y_continuous("") +
  theme_bw()
```

# Analysis

## Marriage constellation, income and happiness

Figures (5)-(7) indicate that reported happiness is associated with labour market affiliation differentially between women and men. In this section we investigate these correlations systematically, particularly the relationship between job-affiliation, family constellation and life-happiness. 

Figure 8 illustrates how reported happiness depends on respondent's income and family constellation. Specifically, we differentiate between the four combinations of marriage and having a high-income job or not. On average, both college educated men and women are more likely to report being very happy when they are married relative to not being married. Furthermore, when not married, both genders report higher happiness when having a high-income job. For married respondents, however, there are considerable gender differences. When married, men continue to report higher happiness when having high-income, whereas the opposite is true for women. This provides preliminary evidence on the thesis that women face difficulties in reconciling having a high-income job and family commitments.

```{r,warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and marriage constellation (college educated sub-sample)"}
z$meanhap_m1 <- NA
z$meanhap_m1[z$married==0 & z$career1==0] <- "Low-inc., unmarried"
z$meanhap_m1[z$married==0 & z$career1==1] <- "High-inc., unmarried"
z$meanhap_m1[z$married==1 & z$career1==0] <- "Low-inc., married"
z$meanhap_m1[z$married==1 & z$career1==1] <- "High-inc., married"

ggplot() + 
  stat_summary(data = z[z$sex == 1 & z$educat==4 ,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4,], aes(x=factor(meanhap_m1), y=vhappyb), 
             fun.y="mean", geom="point", col="Red") +
  scale_x_discrete("Marriage-work combination", limits=c("Low-inc., unmarried", 
              "High-inc., unmarried", "Low-inc., married", "High-inc., married")) +
  scale_y_continuous("Percentage very happy") +
  expand_limits(y=c(20,50)) +
  theme_bw()
```


## Interaction between marriage and job-income

To investigate the gender differences illustrated in figure 8, we deploy different regression models to investigate the correlations further. Table 1 shows the effect of marriage, high-income and the interaction between the two on the binary variable of being very happy. Specification (3)-(4) include no control variables, and the reported coefficient are thus equivalent to the results illustrated in Figure 8. The regression design allows us to deduce that the effect of marriage and having a high-income are significant for both genders at a 10 percent level or lower. Most interestingly, the increase in life-satisfaction from having a high-income job and marriage does not appear to be additive for women: the interaction term has a magnitude of almost minus ten percentage point and it is significant at a 5 percent level. Furthermore, this is not the case for men, where the interaction term of high-income and marriage is practically zero. 

As explained in the research design section, a major methodological challenge is to create a valid counterfactual. Individuals in the different marriage-work constellations are likely to differ from each other in multiple dimensions that correlate with reported happiness. In specification (1) and (2) we control for potential confounding factors that are observable: age, age-squared, survey year, race and cohort (decade of birth). We can observe that the results are robust to their inclusion. 

```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

T1.1 <- lm(vhappyb ~ career1*married, 
         data = subset(z, sex==2 & educat == 4))
T1.2 <- lm(vhappyb ~ career1*married, 
          data = subset(z, sex==1 & educat == 4))
T1.3 <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
         data = subset(z, sex==2 & educat == 4))
T1.4 <- lm(vhappyb ~ career1*married + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
          data = subset(z, sex==1 & educat == 4))

stargazer(T1.3, T1.4, T1.1, T1.2,
    type = 'latex',
    header = FALSE,
    title = "Happiness, marriage and income (college educated men and women)",
    omit.stat = c("f", "rsq", "ser"),
    omit = c("age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men"),
    digits = 2,
    covariate.labels = c("High-income", "Married", "High-income*Married"),
    dep.var.labels   = "Very happy"
    )
```    

## Double-click on married respondent 

A considerable source of sample heterogeneity is likely to arise from unobserved differences between married and unmarried individuals, which affects reported happiness. To address this we circumscribe the sample to married individuals in Table 2. Furthermore, we distinguish between three types of individuals: respondents with high-income, respondents who "keep house", and the residual of low-income, working respondents. The category of individuals who keep house only includes few college educated, married men (~1% of the male sample), but a substantial amount of college educated, married women (~25%). 

Columns (1)-(2) include the controls used in Table 1 as well as the income of the survey respondent’s partner. The spouse’s income is likely to affect both material well-being and the respondent's labour market affiliation, both of which are likely to affect reported happiness.[^pinc_foot] 

[^pinc_foot]: Partner's income is included in the specifications with dummies for each $10.000 income interval.

From specification (2) we observe that married, college educated men report significantly higher happiness when earning high-income relative to low-income, and that men who keep house on average report the lowest happiness, though it is not significantly different from zero. Women, on the other hand, do not report significantly higher happiness when being both married and having a high-income job, and though the coefficient is positive, it is of relatively small magnitude. Women who keep house, however, report significantly higher happiness than women with low-income at the 10 percent level. One explanation for this could be that some women who work part-time would prefer to keep-house, but are not able to for financial reasons.[^spouse_foot] 

[^spouse_foot]: This issue is likely to be mitigated by controlling for the spouse’s income, but could nonetheless prevail as we do not control for region, cost of living, and household wealth.  

Table 2 overall paints a similar picture as Table 1: men are the happiest when they are both married and have a high-income job, whereas this is not the case for women. While there is no direct penalty for having a high-income relative to a low-income job, reported happiness is the highest for married women when keeping house. 

```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

# Create a variable for partner's income
z$othinccat <- trunc(z$def_othinc/10000)
z$othinccat[is.na(z$othinc)] <- -1

# Regressions
T2.1 <- lm(vhappyb ~ career1 + keepinghouse + kid,
           data = subset(z, sex==2 & educat == 4 & married == 1))

T2.2 <- lm(vhappyb ~ career1 + keepinghouse + kid, 
           data = subset(z, sex==1 & educat == 4 & married == 1))

T2.3 <- lm(vhappyb ~ career1 + keepinghouse + kid + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==2 & educat == 4 & married == 1))

T2.4 <- lm(vhappyb ~ career1 + keepinghouse + kid + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==1 & educat == 4 & married == 1))

stargazer(T2.3, T2.4, T2.1, T2.2,
    type = 'latex',
    header = FALSE,
    title = "Happiness and work-status (married, college educated women and men)",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("othinccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Partner's income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men"),
    covariate.labels = c("High-income", "Keeping house", "Child"),
    dep.var.labels   = "Very happy"
   )
```  

## Effect on happiness of young children

```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

# Variables
z$young_child <- 0
z$young_child[z$old1 <= 3 | z$old2 <= 3 | z$old3 <= 3 | z$old4 <= 3 | z$old5 <= 3 | z$old6 <= 3 
              | z$old7 <= 3 | z$old8 <= 3 | z$old9 <= 3 | z$old10 <= 3] <- 1

# Regressions
T3.3 <- lm(vhappyb ~ career1 + keepinghouse + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==2 & educat == 4 & family == 1))

T3.4 <- lm(vhappyb ~ career1 + keepinghouse + as.factor(othinccat) + age + agesq + as.factor(year) +              as.factor(race) + as.factor(bdec), 
           data = subset(z, sex==1 & educat == 4 & family == 1))

T3.5 <- lm(vhappyb ~  keepinghouse*young_child + young_child*career1 + as.factor(othinccat)  + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex==2 & educat == 4 & family == 1))

T3.6 <- lm(vhappyb ~ young_child*career1 + as.factor(othinccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
             data = subset(z, sex==1 & educat == 4 & family == 1))

stargazer(T3.3, T3.4, T3.5, T3.6,
    type = 'latex',
    header = FALSE,
    title = "Happiness and having young children (college educated women and men with a family)",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("othinccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Partner's income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men", "Women", "Men"),
    covariate.labels = c("High-income", "Keeping house", "Young child", 
                         "Keeping House*Young child", "High-income*Young child"),
    dep.var.labels   = "Very happy"
    )
    
```    

One explanation for why women could struggle combining a family and a high-income job relative to men is disproportionate care responsibilities. This is likely to be particularly pronounced when respondents have young children, as they require more intensive care. In Table 3 we thus investigate whether having young children (at least one child below the age of 4) can explain why married women do not report higher levels of happiness when having a high-income job relative to a low-income job. We further restrict the sample to respondents who are both married and have children, such that we compare individuals with young children to individuals with older children. 

Specification (1) and (2) in Table 3 are equivalent to the models in Table 2, except that we only include respondents with children. The results are largely in-line with the results from Table 2, though the effect of having a high-income job for males now only is significant on a 10 percent level.  

Specifications 5 and 6 is augmented with a dummy variable for whether respondents have a young child as well as its interaction with the variables ‘high-income’ and ‘keeping house’. In this specification, we observe that women with a high income, but without a young child, report higher life satisfaction. The coefficient is of high magnitude (approximately ten percentage points) and significant at a 10 percent level. For women with a high-income and a young child, however, the coefficient is negative and of a similar magnitude.  A potential reason for the low significance is the low number of observations for women who have both a high income job and a young child (49 observations). The observation for women is nonetheless interesting as it indicates that women who have children that are four years and older are happier when they have a high-income job, while this seems not to be the case for women with young children. Furthermore, when comparing to the specification for men, we observe no such effects.[^khouse_foot]

[^khouse_foot]: We have excluded the dummy variable for keeping house for males, as the coefficient had an absurdly high (positive) magnitude (approximately 50 percentage points), but remained insignificant due to the negligible amount of observations.  


## Satisfaction with marriage and family constellation. 

Another explanation for why we observe gender differences is that women and men might have different preferences for spouse work characteristics. This is confirmed by Table 4, which shows that the spouse's work status differs substantially between married, college-educated male and female individuals. Regardless of the income, more than 90 percent of female respondents have a partner who works full-time. For male respondents, the picture differs slightly depending on the income group. 54 percent of low-income men have a spouse who works full-time, while this share drops to 42 percent for men with high-income. 

Having a spouse who works less than full-time is likely to alleviate issues of reconciling having a high-income job and family duties, as the spouse is likely to accept disproportionate responsibility for household and care tasks. We analyze this in Table 5 by looking at how the spouse’s work status affects marriage happiness. We use marriage happiness as dependent variable, as tensions over family duties are likely to first and foremost affect quality of family life. In turn, this is likely to be an important component of overall happiness, but when running the regressions of Table 5 on overall happiness we do not find significance at conventional levels. 

```{r, warning=FALSE, echo=FALSE, message=FALSE}

## This code-chunck provides the numbers for the hard-coded latex table used in the paper below.

# Variables

z$sexcat <- NULL
z$sexcat[z$sex==1] <- "Male"
z$sexcat[z$sex==2] <- "Female"

z$spwrkcat <- NULL
z$spwrkcat[z$spwrksta==1] <- "Full-time"
z$spwrkcat[z$spwrksta==2] <- "Part-time"
z$spwrkcat[z$spwrksta==7] <- "Keeping house"


# Gender, income and Spouse Work Status (row probabilities)
ge_sp_ca1 <- table(z$sexcat[z$educat==4 & z$married==1], 
                   z$income_status1[z$educat==4 & z$married==1], 
                   z$spwrkcat[z$educat==4 & z$married==1]) 
# ftable(prop.table(ge_sp_ca1, c(1,2)))
# ftable(ge_sp_ca1)

```


\begin{table}[]
\centering
\caption{Gender, income and spouse work status (row percentages)}
\label{my-label}
\begin{tabular}{llllll}
\hline
       &             & Full-time & Part-time & Keeping house & n    \\ \hline
Female & High-income & 92.45     & 4.15      & 3.4           & 265  \\
       & Low-income  & 95.62     & 0.99      & 3.39          & 1713 \\
Male   & High-income & 42.28     & 33.56     & 24.16         & 745  \\
       & Low-income  & 54.86     & 25.06     & 20.08         & 1285 \\ \hline
\end{tabular}
\end{table}


Specification (1) and (2) of Table 5 is limited to married, college-educated individuals. Women report on average higher marital happiness when their men work full-time rather than part-time (reference category), though the difference is not significant.[^omit]. For men there appear to be no difference between whether the spouse works part- or full-time, and reported happiness is only slightly higher when their spouse stays at home compared to working part-time. 

The low-magnitude effects are expected, as we predominantly expect the family-tensions to exist when respondents have a high-intensity job. In specification (3) and (4) we therefore circumscribe the sample to men and women with a high-income job, which could serve as a proxy for a high-intensity, career job. 

We observe from Table 5 that high-income women report substantially higher marriage happiness when their spouses work full-time. The result, however, remains insignificant, which is likely a result of the small sample size and the small number of high-income women with a partner who works less than full-time. Men, on the other hand, report substantially lower marriage happiness when their partners work full-time, and the coefficient is significant at a 10 percent level. Furthermore, men report higher happiness when their spouse do not work at all relative to working part-time, but it remains insignificant. 

The results of Table 6 indicate two findings. First, that high-income men report higher marital happiness when their partner works less than full-time aligns with the explanation that they face less family tensions over family responsibilities. Second, women seem to prefer a partner who works full-time regardless of income, which was also indicated by Table 5, though one should not put too much weight on the regression results due to the issue of sample size. However, that women have a preference for partners with a similar social and work status has previously been found in studies and could help explain why women face difficulties in reconciling a high-income job with family responsibilities [@saner2015].

[^omit]: We omit the coefficient for women's spouses who stay at home due to too few observations.


```{r, fig.align = "center", results='asis', warning=FALSE, echo=FALSE, message=FALSE}

# Regressions 

T5.1 <- lm(vhapmarb ~ spouse_ft + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex == 2 & educat == 4 & married ==1))

T5.2 <- lm(vhapmarb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
            data = subset(z, sex == 1 & educat == 4 & married ==1))


T5.3 <- lm(vhapmarb ~ spouse_ft + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 2 & educat == 4 & married ==1 & career1 == 1))

T5.4 <- lm(vhapmarb ~ spouse_ft + spouse_home + kid + as.factor(inccat) + age + agesq + as.factor(year) + as.factor(race) + as.factor(bdec), 
               data = subset(z, sex == 1 & educat == 4 & married ==1 & career1 == 1))


stargazer(T5.1, T5.2, T5.3, T5.4,
    type = "latex",
    header = FALSE,
    title = "Marriage happiness and spouse's work-status",
    omit.stat = c("f", "rsq", "ser"),
    digits = 2,
    omit = c("inccat", "age", "agesq", "year", "race", "bdec"),
    omit.labels = c("Family income", "Age", "Age-squared", "Year", "Race", "Cohort"),
    column.labels = c("Women", "Men", "Women", "Men", "Women", "Men"),
    covariate.labels = c("Spouse FT", "Spouse Home", "Children"),
    dep.var.labels   = "Very happy (marriage)",
    add.lines = list(c("Restricted to high-income", "No", "No", "Yes", "Yes"))
    )

```

## Cohorts and norms

In a final step we investigate whether the happiness penalty for women who both have a high-income job and a family persists over generations. Our hypothesis is that younger generations who grew up with more progressive gender norms might find it easier to reconcile family and work. Figure 9 shows differences in average reported happiness for three generations, depending on whether they have a high-income job and/or are married. The oldest cohort is born between 1944 and 1957 (blue), the second cohort between 1958 and 1973 (red) and the youngest cohort between 1974 and 1991 (purple).

Figure 9 indicates that there are generational differences. All generations are considerably happier when being married. Happiness of unmarried women only increases slightly when having a high-income job compared to a low-income job. Among married women, differences between the generations become apparent. While reported happiness decreases for the two older generations when they have a high-income job, happiness slightly increases for the youngest generation. Furthermore, the oldest generation experience the largest reduction in happiness from combining marriage and a high-income job. 

The graphical analysis suggests that "having it all" is more attainable for women in younger cohorts than for their peers from earlier generations. This hints to societal changes that either reflect more progressive gender norms, a more supportive environment that makes it easier for younger generations to combine a high-income job with family responsibilities, or a combination of both. Note, when implemented in a regression framework including controls, we are not able to find statistical significance for the results, which is also a result of very low sample sizes. The results should thus only be interpreted as indicative, and more research is needed to confirm the observation. 


```{r, warning=FALSE,echo=FALSE,cache=TRUE,fig.cap="Happiness and family constellation (college educated women)"}


# Variables
z$meanhap_m1 <- NA
z$meanhap_m1[z$married==0 & z$career1==0] <- "Low-inc., unmarried"
z$meanhap_m1[z$married==0 & z$career1==1] <- "High-inc., unmarried"
z$meanhap_m1[z$married==1 & z$career1==0] <- "Low-inc., married"
z$meanhap_m1[z$married==1 & z$career1==1] <- "High-inc., married"

# ggplot based on three cohorts
ggplot() + 
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 4,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Navyblue") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 5,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Red") +
  stat_summary(data = z[z$sex == 2 & z$educat==4 & z$cohort == 6,], aes(x=factor(meanhap_m1), y=vhappyb), 
               fun.y="mean", geom="point", col="Purple") +
  scale_x_discrete("Marriage-work combination", limits=c("Low-inc., unmarried", "High-inc., unmarried", 
                                                         "Low-inc., married", "High-inc., married")) +
  scale_y_continuous("Percentage very happy") +
  theme_bw()
```

# Conclusion

In our paper we address the research question *"Are career pursuits reconcilable with a happy life?"* This question is motivated by the strive of highly educated women to combine both career and family and the findings of previous studies that this yields in fact no higher life satisfaction. We deployed a four step research strategy to shed light on the potential drivers of this paradox for college educated women.

First, we find that both college-educated men and women are substantially more likely to report being very happy when married. However, the effect on happiness of earning a high-income when married differs between genders. Married men report on average higher life-satisfaction when having a high-income job, while this is not the case for women. Furthermore, married women seem to report higher happiness from ‘keeping house’ compared to women who work but earn a low income, although the difference is only weakly significant.

Second, we find that women with small children appear to report lower happiness when having a high-income job compared to women with older children and high income. This effect does not exist for men. This lends support to the hypothesis that disproportionate distribution of care work is a driver of the happiness penalty that women with high income experience. An alternative explanation would be a conflict between gender roles, i.e. self-perceptions of being a good mother while pursuing a career, which could be amplified in this care intensive time.

Third, our analysis suggests that differences in the spouse’s work status might explain why career women find it harder to reconcile family and job. Women seem happier in a marriage with a full-time worker, while men rather prefer their spouse to either work part-time or stay at home. Being married to a spouse that does not work full-time could alleviate tensions over housework and result in greater reconcilability of job and family for men.

Despite these rather discouraging findings, our descriptive findings suggest that younger generations of women might be less affected by happiness penalties which could point to changing gender norms or a more supportive environment for career women today. Although limitations of the data do not allow us to interpret our findings as causal effects, we believe that the correlations we present hit on some relevant driving factors of gender-specific life satisfaction.



# Software and packages used for the analysis
The analysis is done in R [@CiteR] with the use of the following packages: "ggplot2" [@R-ggplot2], "repmis" [@R-repmis], "plyr" [@R-plyr], "dplyr" [@R-dplyr], "MASS" [@R-MASS], "Hmisc" [@R-Hmisc], "interplot" [@R-interplot], "gridExtra" [@R-gridExtra], "car" [@R-car], "foreign" [@R-foreign], "gmodels" [@R-gmodels], "quantmod" [@R-quantmod] and "reshape" [@R-reshape].

# References